Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-03-02 Nicolas.tex
That's it now

\abstract{The detailed nature of type Ia supernovae (SNe~Ia) remains uncertain,
    and as survey statistics increase, the question of astrophysical systematic
    uncertainties rises, notably that of SN~Ia population evolution. In this
    paper, we study the dependence with redshift of the SN~Ia \texttt{SALT2.4}
    lightcurve stretch, a purely intrinsic SN property, to probe its potential
    redshift drift. The SN stretch has been shown to strongly correlate with the
    SN environment, notably with stellar age tracers. We model the underlying
    stretch distribution as a function of redshift, using the evolution of the
    fraction of young and old SNe~Ia as predicted \textbf{using the SNfactory
    dataset}, and assuming constant underlying stretch distribution for each age
    population consisting of Gaussian mixtures. We test our prediction against
    published samples cut to have marginal magnitude selection effects, so that
    any observed change is indeed astrophysical and not observational in origin.
    In this first study, there are indications that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a better description of the data than any time-constant
    model, including the sample-based asymmetric distributions often used to
    correct Malmquist bias at a significance of over 5 $\sigma$. The favored
    underlying stretch model is \textbf{a bimodal one}, composed of a
    high-stretch mode shared by both young and old environments, and a
    low-stretch mode exclusive to old environments. The precise impact of the
    redshift evolution of the SN Ia population intrinsic properties on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
    distances derived from SN affected by observational selection effects. We
highlight that such a bias will increase with surveys covering increasingly
larger redshift ranges, which is particularly important for LSST.}

@@

The relation between SNe~Ia and their host galaxy properties has been studied
extensively. The first key finding was that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass
host galaxies being brighter on average \cite[e.g.][]{kelly2010, sullivan2010,
childress2013, betoule2014, kim19}. This mass-step correction is currently used
in cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including
for deriving $H_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host galaxies remains unclear, particularly when using
global properties such as the host stellar mass. This in turn raises the
question of the accuracy of such corrections since host galaxy global properties
evolve with redshift. More recently, studies have used the local SN host galaxy
environment to probe more direct connections between the SNe and their host
galaxy environments \citep{rigault2013}, showing that local age tracers such as
the Local specific Star Formation Rate (LsSFR) or the local color are more
strongly correlated with the standardized SN magnitude \citep{roman2018, kim18,
rigault2020}. Further reinforcing this connection is the identification of SN~Ia
spectral features that are correlated with LsSFR \citep{nordin2018}. These
results suggest age as the driving parameter underlying the mass-step. If true,
this would have a significant impact for cosmology, since the environmental
correction to apply to SN standardization could strongly vary with redshift
\citep{rigault2013, childress2014, scolnic2018a}. Yet, the importance of local
SN environmental studies remains highly debated \cite[e.g.][]{jones2015,
jones2019} and especially the impact of such an astrophysical bias has on the
derivation of $H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the younger
(those with a high LsSFR) and which are the older (those with low LsSFR).
Furthermore, since the first SNe~Ia host analyses, the SN stretch has been known
to be strongly correlated with the SN host galaxy properties \citep{hamuy1996,
hamuy2000}. This correlation that has been extensively confirmed since
\citep[e.g.][]{neill2009, lampeitl2010, gupta2011, dandrea2011, pan2014}.
Following the ``A+B'' model and the connection between SN stretch and host
galaxy properties, \cite{howell2007} first discussed the potential redshift
drift of the SN stretch distribution. In this paper we revisit this question
using the most recent SNe~Ia datasets.

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). For this dataset, spectroscopic screening was
    done for candidates with $r \lesssim 19.5$; redshift cuts were then applied
    when selecting which SN Ia to follow, resulting in a redshift range of $0.02
< z < 0.09$, further reduced to $<0.08$ in \cite{rigault2020} for extracting
local host properties. These 114 SNfactory SNe~Ia are thus in the
volume-limited part of the survey (Aldering et al., in prep.), and are therefore
assumed to be a random sampling of the underlying SN population. The SNfactory
sample is particularly useful for studying SN property drift, as it enables us
to have a large complete SN~Ia sample at $z<0.1$.

Finally, we include the HST sample from Pantheon \citep{strolger04}.
These high-redshift SNe are of great interest as they provide the
    greatest leverage for testing evolution. While at these redshifts the
    supernovae typing is challenging, the target classification was robust
    enough to include them within the cosmological analysis \citep{scolnic2018a}
    and we do not impose further cuts. Section~\ref{sec:results} highlights
    that, while compatible with it, our results are not dependent on the
    inclusion of this dataset.

@@

\cite{rigault2020} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia follows
the number of Gyr-old stars in the Universe, i.e. the stellar mass (M$^*$).
Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the fraction of
young (resp.\ old) SNe~Ia in the Universe as a function of redshift, then the
ratio $\delta/\psi$ is expected to follow the evolution of the specific star
formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013, rigault2020, wiseman2020}, in agreement with rate
expectations \citep{mannucci2006, rodney2014}, \cite{rigault2020} concluded that

@@

\bibitem[Saunders et al.(2020)]{saunders2020} Saunders, C., Aldering, G.,
Antilogus, P., et al.\ 2020, VizieR Online Data Catalog, J/ApJ/869/167

@@

>>>>

\abstract{The detailed nature of type Ia supernovae (SNe~Ia) remains uncertain,
    and as survey statistics increase, the question of astrophysical systematic
    uncertainties rises, notably that of SN~Ia population evolution. In this
    paper, we study the dependence with redshift of the SN~Ia \texttt{SALT2.4}
    lightcurve stretch, a purely intrinsic SN property, to probe its potential
    redshift drift. The SN stretch has been shown to strongly correlate with the
    SN environment, notably with stellar age tracers. We model the underlying
    stretch distribution as a function of redshift, using the evolution of the
    fraction of young and old SNe~Ia as predicted using the SNfactory dataset,
    and assuming constant underlying stretch distribution for each age
    population consisting of Gaussian mixtures. We test our prediction against
    published samples cut to have marginal magnitude selection effects, so that
    any observed change is indeed astrophysical and not observational in origin.
    In this first study, there are indications that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a better description of the data than any time-constant
    model, including the sample-based asymmetric distributions often used to
    correct Malmquist bias at a significance of over 5 $\sigma$. The favored
    underlying stretch model is a bimodal one, composed of a high-stretch mode
    shared by both young and old environments, and a low-stretch mode exclusive
    to old environments. The precise impact of the redshift evolution of the SN
    Ia population intrinsic properties on cosmology remains to be studied. Yet,
    the astrophysical drift of the SN stretch distribution does affect current
    Malmquist bias corrections and thereby distances derived from SN affected by
    observational selection effects. We highlight that such a bias will increase
with surveys covering increasingly larger redshift ranges, which is particularly
important for LSST.}

@@

The relation between SNe~Ia and their host galaxy properties has been studied
extensively. The first key finding was that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass
host galaxies being brighter on average \cite[e.g.][]{kelly2010, sullivan2010,
childress2013, betoule2014, kim19, smith2020}. This mass-step correction is
currently used in cosmological analyses \citep[e.g.][]{betoule2014,
scolnic2018a}, including for deriving $H_0$ \citep{riess2016, riess2019}. Yet,
the underlying connection between the SNe and their host galaxies remains
unclear, particularly when using global properties such as the host stellar
mass. This in turn raises the question of the accuracy of such corrections since
host galaxy global properties evolve with redshift. More recently, studies have
used the local SN host galaxy environment to probe more direct connections
between the SNe and their host galaxy environments \citep{rigault2013}, showing
that local age tracers such as the Local specific Star Formation Rate (LsSFR) or
the local color are more strongly correlated with the standardized SN magnitude
\citep{roman2018, kim18, rigault2020}. Further reinforcing this connection is
the identification of SN~Ia spectral features that are correlated with LsSFR
\citep{nordin2018}. These results suggest age as the driving parameter
underlying the mass-step. If true, this would have a significant impact for
cosmology, since the environmental correction to apply to SN standardization
could strongly vary with redshift \citep{rigault2013, childress2014,
scolnic2018a}. Yet, the importance of local SN environmental studies remains
highly debated \cite[e.g.][]{jones2015, jones2019} and especially the impact of
such an astrophysical bias has on the derivation of $H_0$ \citep{jones2015,
riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, smith2012} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the younger
(those with a high LsSFR) and which are the older (those with low LsSFR).
Furthermore, since the first SNe~Ia host analyses, the SN stretch has been known
to be strongly correlated with the SN host galaxy properties \citep{hamuy1996,
hamuy2000}. This correlation that has been extensively confirmed since
\citep[e.g.][]{neill2009, lampeitl2010, gupta2011, dandrea2011, pan2014}.
Following the ``A+B'' model and the connection between SN stretch and host
galaxy properties, \cite{howell2007} first discussed the potential redshift
drift of the SN stretch distribution. In this paper we revisit this question
using the most recent SNe~Ia datasets.

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2 ; see \citealt{aldering2020}). For this dataset, spectroscopic
screening was done for candidates with $r \lesssim 19.5$; redshift cuts were
then applied when selecting which SN Ia to follow, resulting in a redshift range
of $0.02 < z < 0.09$, further reduced to $<0.08$ in \cite{rigault2020} for
extracting local host properties. These 114 SNfactory SNe~Ia are thus in the
volume-limited part of the survey (Aldering et al., in prep.), and are therefore
assumed to be a random sampling of the underlying SN population. The SNfactory
sample is particularly useful for studying SN property drift, as it enables us
to have a large complete SN~Ia sample at $z<0.1$. Finally, we include the HST
sample from Pantheon \citep{strolger04}. These high-redshift SNe are of great
interest as they provide the greatest leverage for testing evolution. While at
these redshifts the supernovae typing is challenging, the target classification
was robust enough to include them within the cosmological analysis
\citep{scolnic2018a} and we do not impose further cuts.
Section~\ref{sec:results} highlights that, while compatible with it, our results
are not dependent on the inclusion of this dataset.

@@

\cite{rigault2020} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, smith2012, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia follows
the number of Gyr-old stars in the Universe, i.e. the stellar mass (M$^*$).
Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the fraction of
young (resp.\ old) SNe~Ia in the Universe as a function of redshift, then the
ratio $\delta/\psi$ is expected to follow the evolution of the specific star
formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013, rigault2020, wiseman2020}, in agreement with rate
expectations \citep{mannucci2006, rodney2014}, \cite{rigault2020} concluded that

@@

\bibitem[Saunders et al.(2018)]{saunders2018} Saunders, C., Aldering, G.,
Antilogus, P., et al.\ 2018, \apj, 869, 167. doi:10.3847/1538-4357/aaec7e

@@

\bibitem[Smith et al.(2012)]{smith2012} Smith, M., Nichol, R.~C., Dilday, B., et
al.\ 2012, \apj, 755, 61

\bibitem[Smith et al.(2020)]{smith2020} Smith, M., Sullivan, M., Wiseman, P., et
al.\ 2020, \mnras, 494, 4426
